---
title: 'Introduction to Bioconductor'
subtitle: '<i class="fab fa-r-project"></i> for genomics data science (CCB cluster)'
author: "Kevin Rue-Albrecht"
institute: "Oxford Biomedical Data Science Training Programme"
date: "2023-02-08 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, metropolis, rladies-fonts, "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
# uncomment this line to produce HTML and PDF in RStudio:
knit: pagedown::chrome_print
---

```{r, child='goals_and_objectives.Rmd'}
```

---

# Prerequisites

- An account on the teaching cluster.

- A [<i class="fab fa-github"></i> GitHub](https://github.com/) repository for this course.

## Lessons

- Introduction to base <i class="fab fa-r-project"></i>.

- Object Oriented Programming (OOP).

- Introduction to `r BiocStyle::CRANpkg("ggplot2")`.

---

# Set up

- Launch RStudio Server on the teaching cluster.

- Start a new RStudio project called `bioconductor`.

- Make a copy of the template notebook for this lesson in your <i class="fab fa-git"></i> repository.

- Make symbolic link from the RStudio project to the template notebook.

- Open the notebook and follow along, editing and running the code as needed.

```{r, include=TRUE, echo=FALSE, fig.align='center', out.height='250px'}
## Source: Adobe Illustrator (Kevin Rue-Albrecht)
knitr::include_graphics("img/obds-git-setup.svg")
```

---

```{r, child='philosophy.Rmd'}
```

---

# Exercise

## BiocManager

On the teaching cluster, all the packages required for the course are already installed.
As a result, you do not need (and should not) install any package yourself.

> Instead, you should ask the system administrators to install or update packages for you.

To give you an idea, what the system administrators did essentially comes down to:

```{r, include=TRUE, echo=TRUE, eval=FALSE}
install.packages("BiocManager")
BiocManager::install(c("Biostrings", "S4Vectors"))
```

The package `r BiocStyle::CRANpkg("BiocManager")` is the only official Bioconductor package on CRAN.
It provides functions to install, update, and more generally manage your library of Bioconductor (and CRAN) packages.

The function `BiocManager::install()` can install packages from Bioconductor, CRAN, and even GitHub.
Here, we demonstrate it with a vector of two package names.
However, it can be used to install a single package name as a character value.

- Run the function `BiocManager::valid()`.
  What does it do?

---

```{r, child='etc.Rmd'}
```
