---
title: 'Introduction to Bioconductor'
subtitle: '<i class="fab fa-r-project"></i> for genomics data science'
author: "Kevin Rue-Albrecht"
institute: "Oxford Biomedical Data Science Training Programme"
date: "2022-10-19 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, metropolis, rladies-fonts, "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
# uncomment this line to produce HTML and PDF in RStudio:
knit: pagedown::chrome_print
---

```{r, child='goals_and_objectives.Rmd'}
```

---

# Prerequisites

- An [RStudio Cloud](https://rstudio.cloud/) account.

- A clone of the shared GitHub repository for this course.

## Lessons

- Introduction to base <i class="fab fa-r-project"></i>.

- Introduction to `r BiocStyle::CRANpkg("ggplot2")`.

---

# Set up

- Create a directory named `r_bioconductor`.

- Make a copy of the file `template.Rmd` for this lesson from the Git repository to the new directory.

- Open your copy of the template notebook for this lesson in RStudio.

- Follow along, running code from the notebook as mentioned during the lesson.

---

```{r, child='etc.Rmd'}
```
