---
title: 'Introduction to Bioconductor'
subtitle: '<i class="fab fa-r-project"></i> for genomics data science'
author: "Kevin Rue-Albrecht"
institute: "University of Oxford (for the African Institute of Biomedical Science and Technology)"
date: "2022-11-16 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, metropolis, rladies-fonts, "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
# uncomment this line to produce HTML and PDF in RStudio:
knit: pagedown::chrome_print
---

```{r, child='goals_and_objectives.Rmd'}
```

---

# Prerequisites

- A computer with [Microsoft Windows](https://www.microsoft.com/en-gb/windows?r=1).

- A working installation of [<i class="fab fa-r-project"></i>](https://cran.r-project.org/).

- A working installation of [RStudio Desktop](https://posit.co/download/rstudio-desktop/).

## Lessons

- Introduction to base <i class="fab fa-r-project"></i>.

- Object Oriented Programming (OOP).

- Introduction to `r BiocStyle::CRANpkg("ggplot2")`.

---

# Set up

- Launch RStudio Desktop on your Windows computer.

- Make a copy of the template notebook for this lesson in your <i class="fab fa-git"></i> repository.

- Make a symbolic link to your copy of the notebook in the RStudio project for this week.

- Open the notebook and follow along, editing and running the code as needed.

---

```{r, child='philosophy.Rmd'}
```

---

# Exercise

## BiocManager

Prior to the course, we have provided you with an R script that installed
all the packages required for the course.
As a result, you do not need install any package during this course.

Briefly, what the R script does essentially comes down to:

```{r, include=TRUE, echo=TRUE, eval=FALSE}
install.packages("BiocManager")
BiocManager::install(c("Biostrings", "S4Vectors"))
```

The package `r BiocStyle::CRANpkg("BiocManager")` is the only official Bioconductor package on CRAN.
It provides functions to install, update, and more generally manage your library of Bioconductor (and CRAN) packages.

The function `BiocManager::install()` can install packages from Bioconductor, CRAN, and even GitHub.
Here, we demonstrate it with a vector of two package names.
However, it can be used to install a single package name as a character value.

- Run the function `BiocManager::valid()`.
  What does it do?

---

```{r, child='etc.Rmd'}
```
