---
title: 'Introduction to Bioconductor'
subtitle: '<i class="fab fa-r-project"></i> for data science and genomics'
author: "Devika Agarwal & Kevin Rue"
institute: "Oxford Biomedical Data Science Training Programme"
date: "2025-05-20 (updated: `r Sys.Date()`)"
date-format: long
format:
  revealjs: 
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    logo: img/ox_brand1_pos.gif
    css: styles.css
    fontsize: 1.5em
    footer: Devika Agarwal & Kevin Rue -- Introduction to Bioconductor
# run the following line after knitting to convert HTML to PDF
# renderthis::to_pdf("index.cluster.html")
---

```{r setup, include = FALSE}
stopifnot(requireNamespace("htmltools"))
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
options(width = 90)
stopifnot(require(tidyverse))
stopifnot(require(S4Vectors))
```

```{r, load_refs, include=FALSE, cache=FALSE}
options(htmltools.dir.version = FALSE)
library(RefManageR)
BibOptions(
  check.entries = FALSE,
  bib.style = "authoryear",
  cite.style = "authoryear",
  max.names = 2,
  style = "markdown",
  hyperlink = "to.doc",
  dashed = TRUE)
bib <- ReadBib("bibliography.bib")
```

```{r}
#| child: goals_and_objectives.qmd
```

## Credits

- Liezel Tamon
- Kevin Rue-Albrecht

## Prerequisites

- An account on the teaching cluster.

### Lessons

- Introduction to base <i class="fab fa-r-project"></i>.
- Object Oriented Programming (OOP).
- Introduction to `r BiocStyle::CRANpkg("ggplot2")`.

## Set up

- Launch RStudio Server on the teaching cluster.
- Start a new RStudio project called `bioconductor`.
- Make a copy of the template notebook for this lesson in your RStudio project.
- Open the notebook and follow along, editing and running the code as needed.

```{r}
#| child: philosophy.qmd
```

## Exercise

### BiocManager

On the teaching cluster, all the packages required for the course are already installed.
As a result, you do not need (and should not) install any package yourself.

Instead, you should ask the system administrators to install or update packages for you.

To give you an idea, what the system administrators did essentially comes down to:

```{r}
#| echo: true
#| eval: false
install.packages("BiocManager")
BiocManager::install(c("Biostrings", "S4Vectors"))
```

The package `r BiocStyle::CRANpkg("BiocManager")` is the only official Bioconductor package on CRAN.
It provides functions to install, update, and more generally manage your library of Bioconductor (and CRAN) packages.

The function `BiocManager::install()` can install packages from Bioconductor, CRAN, and even GitHub.
Here, we demonstrate it with a vector of two package names.
However, it can be used to install a single package name as a character value.

- Run the function `BiocManager::valid()`.
  What does it do?

```{r}
#| child: etc.qmd
```
