---
title: "Example code for introduction to Bioconductor"
author: "Kevin Rue-Albrecht"
date: "23/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Biostrings)
library(BSgenome)
library(BSgenome.Hsapiens.UCSC.hg38.masked)
library(rtracklayer)
library(org.Hs.eg.db)
library(biomaRt)
library(SummarizedExperiment)
library(tximeta)
library(tximportData)
library(DESeq2)
library(SingleCellExperiment)
library(ExperimentHub)
library(iSEE)
library(scuttle)
library(scater)
```

# Exercise

## Exploring software packages

- Visit the listing of packages on the Bioconductor [biocViews](https://bioconductor.org/packages/release/BiocViews.html) web page.

- Use the `Autocomplete biocViews search` box in the upper left to identify packages that have been tagged for RNA sequencing analysis.

- Explore other analysis like ChIP-seq, epigenetics, variant annotation, proteomics, single-cell genomics, etc.

- Explore the graph of software packages by expanding and contracting individual terms.

- In the RNA-seq category, find out which of `DESeq2` and `edgeR` is more popular, and go to their landing page.

- Briefly explore the vignette and reference manual links.

When would you consult the vignette? When would the reference manual be helpful?

> Answer:
> The vignette is great at showcasing the set of functions in a package
> in the context of a typical workflow. This is much most convenient for new users
> and users who wish to refresh their knowledge of the package and best practices.
> 
> In contrast, the manual simply collates the manual pages of the package.
> This is most useful for users who wish to check the arguments of any given function
> in the package, without having to launch an R session and access the help page itself.

# Demo

## Bioconductor uses the S4 class system

Define the class.

```{r}
setClass("Person",
  representation(
    name = "character",
    age = "numeric")
  )
```

Define the constructor function.

```{r}
Person <- function(name, age) {
  new("Person", name = name, age = age)
}
```

Use the constructor function to make an object.

```{r}
kevin <- Person(name = "Kevin", age = 21)
kevin
```

Don't access internal slots directly.

```{r}
kevin@age # don't do that
```

The package developers should provide you with accessor functions.

```{r}
get_age <- function(object) {
  object@age
}
```

In which case, you should use them.

```{r}
get_age(kevin) # do that (if possible)
```

# Exercise

## Install Bioconductor

On the teaching cluster, all the packages required for the course are already installed.
As a result, you do not need (and should not) install any package yourself.

> Instead, you should ask the system administrators to install or update packages for you.

To give you an idea, what the system administrators did essentially comes down to:

```{r, include=TRUE, echo=TRUE, eval=FALSE}
install.packages("BiocManager")
BiocManager::install(c("Biostrings", "S4Vectors"))
```

The package `r BiocStyle::CRANpkg("BiocManager")` is the only official Bioconductor package on CRAN.
It provides functions to install, update, and more generally manage your library of Bioconductor (and CRAN) packages.

The function `BiocManager::install()` can install packages from Bioconductor, CRAN, and even GitHub.
Here, we demonstrate it with a vector of two package names.
However, it can be used to install a single package name as a character value.

- Run the function `BiocManager::valid()`.
  What does it do?

```{r}
BiocManager::valid()
```

> Answer:
> It checks if all the Bioconductor packages in the library are up-to-date:
> 
> - If yes, it returns `TRUE`.
> - If no, it prints the command that should be run to update the library.

# Exercise

## Use Bioconductor packages and help pages

- Load the `Biostrings` package.

```{r}
library(Biostrings)
```

- The package provides the object `DNA_ALPHABET`.
  Print the object in the console and use the help page to explain the contents of the object.

```{r}
DNA_ALPHABET
```

```{r, eval=FALSE}
help(DNA_ALPHABET)
```

- The package provides the function `IlluminaQuality()`.
  Run some code from the "Examples" section of the help page, and describe what the function does.
  
```{r, eval=FALSE}
help(IlluminaQuality)
```
  
```{r}
IlluminaQuality(0:40)
```

- Access the package vignette(s).
  How many vignette does the package include?

```{r, eval=FALSE}
browseVignettes("Biostrings")
```

- Import sequences in the file `TruSeq3-PE-2.fa`.
  What is the class of the object?

```{r}
truseq_adapters <- Biostrings::readAAStringSet("data/TruSeq3-PE-2.fa")
truseq_adapters
```

# Exercise

## BSgenome packages

- Load the package `BSgenome`.

```{r}
library(BSgenome)
```

- Use the function `BSgenome::available.genomes()`.
  What does it do?

```{r}
BSgenome::available.genomes() %>% head()
```

- Load the package `BSgenome.Hsapiens.UCSC.hg38.masked`.

```{r}
library(BSgenome.Hsapiens.UCSC.hg38.masked)
```

- Assign the genome object provided in the package to a new object called `genome`.

```{r}
genome <- BSgenome.Hsapiens.UCSC.hg38.masked
genome
```

An E-box (enhancer box) is a DNA response element found in some eukaryotes that acts as a protein-binding site and has been found to regulate gene expression in neurons, muscles, and other tissues.

The E-box motif is `"CANNTG"`.

- Extract the sequence of chromosome Y from the `genome` object above.

```{r}
chrY <- genome$chrY
```

- Count the number of E-box motifs present on chromosome Y using the appropriate method available in the `Biostrings` package.

```{r}
Ebox <- "CANNTG"
countPattern(Ebox, chrY, fixed=FALSE)
```

# Demo

## Creating genomic ranges

```{r}
demo_granges <- GRanges(
  seqnames = c("chr1", "chr2"),
  ranges = IRanges(
    start = c(10, 20),
    end = c(25, 35)),
  metadata1 = c("control", "target"),
  metadata2 = c(1, 2))
demo_granges
```

**What genomics file format does this remind you of?**

> Answer:
> - BED or GTF file: genomic coordinates, accompanied with metadata.

# Demo

## Accessing the contents of GRanges objects

```{r}
seqnames(demo_granges)
```

```{r}
start(demo_granges)
```

```{r}
end(demo_granges)
```

```{r}
seqlevels(demo_granges)
```

# Demo

## Metadata columns

```{r}
mcols(demo_granges)
```

```{r}
demo_granges$metadata1
```

```{r}
metadata(demo_granges)
```

# Demo

## DataFrame

```{r}
library(S4Vectors)
demo_DataFrame <- DataFrame(
    column1 = c("A", "B"),
    column2 = c(1, 2)
)
demo_DataFrame
```

```{r}
mcols(demo_DataFrame) <- DataFrame(
  column_type = c("character", "numeric")
)
mcols(demo_DataFrame)
```

# Exercise

## rtracklayer

- Load the `rtracklayer` package.

```{r}
library(rtracklayer)
```

- Import the file `Homo_sapiens.GRCh38.104.gtf.gz` using the function `rtracklayer::import()`.
  Assign value to an object called `gtf_data`.

```{r}
gtf_data <- rtracklayer::import("data/Homo_sapiens.GRCh38.104.gtf.gz")
gtf_data
```

- What are the metadata columns available in the object?

```{r}
colnames(mcols(gtf_data))
```

- Use the function `subset()` to filter the annotations for the gene `ACTB`.

```{r}
gtf_ACTB <- subset(gtf_data, gene_name == "ACTB")
nrow(gtf_ACTB)
```

- How many transcripts do the annotations describe for that gene?

```{r}
gtf_ACTB$transcript_id %>% 
    unique() %>% 
    setdiff(NA) %>% 
    length()
```

- Subset the annotations to exons, and use the `split()` function to separate exons from different transcripts.
  What is the class of the object that is returned?

```{r}
gtf_ACTB_exons <- subset(gtf_ACTB, type == "exon")
gtf_ACTB_transcripts <- split(gtf_ACTB_exons, gtf_ACTB_exons$transcript_id)
gtf_ACTB_transcripts
```

# Exercise

## Annotation packages

Each annotation packages contains a database of information.

The `org.db` family of packages provide annotations that link  several types of identifiers for genes, proteins, and pathways (e.g. KEGG, Gene Ontology).

- Load the package `org.Hs.eg.db`.

```{r}
library(org.Hs.eg.db)
```

Each `OrgDb` package contains an object of the same name as the package.

- What is the class of the object?

```{r}
class(org.Hs.eg.db)
```

- Use the `columns()` method to discover which sorts of annotations can be extracted
from the object.

```{r, include=TRUE}
columns(org.Hs.eg.db)
```

- Use the `keytypes()` method to discover which columns can be used to query information.

```{r}
keytypes(org.Hs.eg.db)
```

- Use the `keys()` method to get the first six gene symbols in the database.
  Store as `human_symbols`.

```{r}
human_symbols <- keys(org.Hs.eg.db, "SYMBOL") %>% 
  head()
human_symbols
```

- Use the `AnnotationDbi::select()` method to fetch the Entrez and Ensembl identifiers for those gene symbols.

```{r}
AnnotationDbi::select(org.Hs.eg.db, human_symbols, c("ENTREZID", "ENSEMBL"), "SYMBOL")
```

  + What happens if you do not specify `AnnotationDbi`?

> Answer:
> - Depending on the order in which you attached packages to your R session,
> you may see the error message:
> ```
> no applicable method for 'select' applied to an object of class
> "c('OrgDb', 'AnnotationDb', 'envRefClass', '.environment', 'refClass', 'environment', 'refObject', 'AssayData')"
> ```

- Use the `mapIds()` method to get the Ensembl identifier for those gene symbols.

```{r}
mapIds(org.Hs.eg.db, human_symbols, "ENSEMBL", "SYMBOL")
```

# Exercise

## biomaRt

- Load the package `biomaRt`.

```{r}
library(biomaRt)
```

- Use the `listMarts()` function to list the BioMart databases to which the `biomaRt` package can connect.

```{r}
listMarts()
```

- Use the `listEnsemblArchives()` function to lists the available archived versions of Ensembl.

```{r}
listEnsemblArchives()
```

- Use the `useMart()` function to create an object named `mart` using the `"ENSEMBL_MART_ENSEMBL"` BioMart database and set the host option to `"https://may2021.archive.ensembl.org"`.
  Why would one do that in practice?

```{r}
mart <- useMart("ENSEMBL_MART_ENSEMBL", host = "https://may2021.archive.ensembl.org")
mart
```

> Answer:
> - In practice, users choose a host for reproducibility.
> - Otherwise, the latest information is returned, which may change without notice.

- Use the `listDatasets()` function to list the datasets available in the selected BioMart database.
  What is the identifier of the database that contains information for _Homo sapiens_?

```{r}
listDatasets(mart) %>% head()
```

- Use again the `useMart()` function, and replace the `mart` object by a new one that points to information for _Homo sapiens_.

```{r}
mart <- useMart(
  biomart = "ENSEMBL_MART_ENSEMBL",
  dataset = "hsapiens_gene_ensembl",
  host = "https://may2021.archive.ensembl.org"
)
mart
```

- Use the `listAttributes()` function to list the fields of information available in the dataset.

```{r}
listAttributes(mart) %>% head()
```

- Use the `getBM()` function to fetch the chromosome, start and end positions, and strand information for the gene symbols that you stored earlier as `human_symbols`.

```{r}
getBM(
  attributes = c("hgnc_symbol", "chromosome_name", "start_position", "end_position", "strand"),
  filters = "hgnc_symbol",
  values = human_symbols,
  mart = mart
)
```

# Demo

## EnsDb packages

```{r}
library(EnsDb.Hsapiens.v86)
```

```{r}
class(EnsDb.Hsapiens.v86)
```

```{r}
columns(EnsDb.Hsapiens.v86)
```

# Demo

## TxDb packages

```{r}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
class(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
columns(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
keytypes(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
keys(TxDb.Hsapiens.UCSC.hg38.knownGene, "GENEID") %>% head()
```

# Exercise

## Creating a SummarizedExperiment object

```{r, include=TRUE, results='hide'}
library(SummarizedExperiment)
demo_count_matrix <- matrix(
  data = c(1, 2, 3, 4, 5, 6, 7, 8), nrow = 4, ncol = 2,
  dimnames = list(
    paste0("gene", 1:4),
    paste0("sample", 1:2)
  ))
demo_rowdata <- DataFrame(
  symbol = head(letters, 4),
  length = c(1234, 5678)
  )
demo_coldata <- DataFrame(
  condition = c("control", "treated"),
  age = c(18, 20)
  )
demo_se <- SummarizedExperiment(
  assays = list(
    counts = demo_count_matrix),
  rowData = demo_rowdata,
  colData = demo_coldata)
```

```{r, include=TRUE}
demo_se
```

Use the methods `assays()`, `assayNames()` `assay()`, `colData()`, `rowData()`, `mcols()` and `metadata()` on the object `demo_se`.

```{r}
assays(demo_se)
assayNames(demo_se)
assay(demo_se, "counts")
colData(demo_se)
rowData(demo_se)
mcols(demo_se)
metadata(demo_se)
```

What do those functions do?

> Answer:
> - Those are accessor functions.
> - Each of them returns a component of the object.

# Exercise

## Assembling a SummarizedExperiment object

- Import the matrix in the file `counts.csv`.
  Call it `assay_counts`.

```{r}
assay_counts <- read.csv("data/summarized_experiment/counts.csv", row.names = 1)
assay_counts <- as.matrix(assay_counts)
head(assay_counts)
```

- Import the data frames in the files `sample_metadata.csv` and `gene_metadata.csv`.
  Call them `sample_metadata` and `gene_metadata`.

```{r}
sample_metadata <- read.csv("data/summarized_experiment/sample_metadata.csv", row.names = 1)
head(sample_metadata)
```

```{r}
gene_metadata <- read.csv("data/summarized_experiment/gene_metadata.csv", row.names = 1)
head(gene_metadata)
```

- Use the function `SummarizedExperiment()` to combine those three objects into a single `SummarizedExperiment` object.
  Call it `demo_se`.
  
  + Assign the matrix to an assay named `counts`.

```{r}
demo_se <- SummarizedExperiment(
  assays = list(counts = assay_counts),
  colData = sample_metadata,
  rowData = gene_metadata
)
demo_se
```

# Demo

## Inheritance and coercion

```{r}
demo_dataframe <- data.frame(A = c(1, 2), B = c(3, 4))
demo_dataframe
```

```{r}
demo_DataFrame <- as(demo_dataframe, "DataFrame")
demo_DataFrame
```

# Exercise

## SummarizedExperiment extensions

- Load the package `DESeq2`.

```{r}
library(DESeq2)
```

- Convert your earlier `demo_se` to a `DESeqDataSet` object, using the `as()` function.
  Call the object `demo_deseq`.

```{r}
demo_deseq <- as(demo_se, "DESeqDataSet")
demo_deseq
```

- What are the slots of information present in the `DESeqDataSet` object?
  Use the function `slotNames()`.

```{r}
slotNames(demo_deseq)
```

- Which slots are new compared to the `SummarizedExperiment` object?

```{r}
slotNames(demo_se)
```

- Which components of the `DESeqDataSet` object can you access using accessor functions (e.g., `assays()`) ?
  Do all of those accessors work on the original `demo_se` object?

```{r}
assays(demo_se)
assayNames(demo_se)
assay(demo_se, "counts")
colData(demo_se)
rowData(demo_se)
mcols(demo_se)
metadata(demo_se)
design(demo_deseq)
dispersionFunction(demo_deseq)
# design(demo_se) # error
# dispersionFunction(demo_se) # error
```

> Answer:
> - The methods `design()` and `dispersionFunction()` only work on `DESeqDataSet` objects.
> - They do not work on the original `SummarizedExperiment` object.

# Exercise

## SingleCellExperiment

The class name `SingleCellExperiment` is only a name.
While developed with single-cell genomics in mind, it is not limited to that use.
Remember that classes are simply defined by their slots and methods.
If the slots and methods are suitable to store and process another type of data, then go for it!

- Install and load the package `SingleCellExperiment`.

```{r}
library(SingleCellExperiment)
```

- Convert your earlier `demo_se` to a `SingleCellExperiment` object, using the `as()` function.
  Call the object `demo_sce`.

```{r}
demo_sce <- as(demo_se, "SingleCellExperiment")
```

- Display the object.
  What additional information can the `SingleCellExperiment` class store, relative to the `SummarizedExperiment` class?

```{r}
demo_sce
```

> Answer:
> Additional information include `reducedDimNames`, `mainExpName`, and `altExpNames`.

- Use the methods `reducedDimNames()`, `reducedDims()`, and `reducedDim()` on the object `demo_sce`.
  What do they do?

```{r}
reducedDimNames(demo_sce)
reducedDims(demo_sce)
# reducedDim(demo_sce) # error
```

> Answer:
> - `reducedDimNames()` displays the names of dimensionality reduction results available.
> - `reducedDims()` returns the list of dimensionality reduction results available.
> - `reducedDim()` returns a single dimensionality reduction result, by default the first one, if any. If there are none, it throws an error.

- **Bonus point:** Use the method `reducedDim()` to store a new dimensionality reduction matrix called `"PCA"` in the object.
  Display and inspect the updated object.

```{r}
pca_object <- prcomp(t(assay(demo_sce)))
reducedDim(demo_sce, "PCA") <- pca_object$x
demo_sce
```

# Demo

## The ExperimentHub package and resources

```{r, include=TRUE}
library(ExperimentHub)
ehub <- ExperimentHub()
ehub
```

```{r}
query(ehub, c("SingleCellExperiment"))
```

```{r, include=TRUE}
eh1433 <- ehub[["EH1433"]]
eh1433
```

# Demo

## Description of a dataset in the ExperimentHub

```{r}
ehub["EH1433"]
```

# Exercise

## Interactive visualisation using iSEE

- Load the package `iSEE`.

```{r}
library(iSEE)
```

- Use the `iSEE()` function on the `SingleCellExperiment` object `eh1433`.

```{r, include=TRUE, eval=FALSE}
iSEE(eh1433)
```

- Subset the object `eh1433` to remove cells where the sum of counts is `NA`.

```{r}
eh1433 <- eh1433[, !is.na(colSums(assay(eh1433)))]
eh1433
```

- Use the packages `scuttle` and `scater` to compute and add a PCA result to the object `eh1433`

```{r}
library(scuttle)
library(scater)
eh1433 <- logNormCounts(eh1433)
eh1433 <- runPCA(eh1433)
```

- Launch the `iSEE` GUI again.
  What changed?

```{r, eval=FALSE}
iSEE(eh1433)
```

# Advanced

## BiocManager repositories

- Run `options("repos")`.
  What do you see?

```{r}
options("repos")
```

The function `options()` can be used to get or set global options of the current <i class="fab fa-r-project"></i> session.

- Run `BiocManager::repositories()`.

```{r}
BiocManager::repositories()
```

The function reports the URLs from which to install Bioconductor and CRAN packages.

- Set the `repos` option to the value of `BiocManager::repositories()`.

```{r}
options(repos = BiocManager::repositories())
```

- Run `options("repos")` again.
  What do you see now?
  What does it mean?

```{r}
options("repos")
```

> Answer:
> Now, the `repos` option is set to the value of `BiocManager::repositories()`.
> It means that even the base R function `install.packages()` will search for packages
> on the Bioconductor repositories, not only CRAN.

# Advanced

## Import SummarizedExperiment objects using tximeta

- Load the packages `tximeta` and `tximportData`.

```{r}
library(tximeta)
```

The help page of the `tximeta()` function refers to an example output of the Salmon program that is installed on your computer at `system.file("extdata/salmon_dm/SRR1197474/quant.sf", package="tximportData")`.

```{r, eval=FALSE}
help(tximeta)
```

- Navigate to that location in the file explorer of your respective operating system, and examine its contents, in a text editor or your terminal application.

```{r, eval=FALSE}
file.edit(system.file("extdata/salmon_dm/SRR1197474/quant.sf", package="tximportData"))
```

- Run the rest of the example code, and discuss the use of a linked transcriptome with `tximeta` (read the help page of `makeLinkedTxome()` and set `write=TRUE` instead of using the example code as-is).

```{r}
dir <- system.file("extdata/salmon_dm", package="tximportData")
files <- file.path(dir, "SRR1197474", "quant.sf") 
coldata <- data.frame(files, names="SRR1197474", condition="A", stringsAsFactors=FALSE)
```

```{r}
se <- tximeta(coldata)
se
```

```{r}
indexDir <- file.path(dir, "Dm.BDGP6.22.98_salmon-0.14.1")
fastaFTP <- c("ftp://ftp.ensembl.org/pub/release-98/fasta/drosophila_melanogaster/cdna/Drosophila_melanogaster.BDGP6.22.cdna.all.fa.gz",
              "ftp://ftp.ensembl.org/pub/release-98/fasta/drosophila_melanogaster/ncrna/Drosophila_melanogaster.BDGP6.22.ncrna.fa.gz")
gtfPath <- file.path(dir, "Drosophila_melanogaster.BDGP6.22.98.gtf.gz")
makeLinkedTxome(indexDir=indexDir, source="Ensembl", organism="Drosophila melanogaster",
                release="98", genome="BDGP6.22", fasta=fastaFTP, gtf=gtfPath, write=TRUE, jsonFile = "tximeta.json")
se <- tximeta(coldata)
se
```

- What other types of quantification programs does `tximeta` support?
  Where do you find that information?

```{r, eval=FALSE}
help(tximport)
```

> Answer:
> In the help page of the function `tximport()`.
