---
title: 'Object Oriented Programming (OOP)'
subtitle: '<i class="fab fa-r-project"></i>'
author: "Kevin Rue-Albrecht"
institute: "Oxford Biomedical Data Science Training Programme"
date: "2022-01-27 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, metropolis, rladies-fonts, "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
# uncomment this line to produce HTML and PDF in RStudio:
knit: pagedown::chrome_print
---

```{r, child='opening.Rmd'}
```

---

# Prerequisites

- An account on the CCB cluster.

- A clone of the shared GitHub repository for this course.

---

# Set up

- Checkout and pull the main branch of the shared repository.

- Create and launch a new RStudio project in `/project/obds/$USER/projects/object_oriented_programming_ggplot2`.

- Make a copy of the file `template.Rmd` for this lesson from the Git repository to your project.
  Use your username in the filename to make it unique, e.g.

.small-code[

> ```
> cp \
>   /project/obds/$USER/git/week2day4-object_oriented_programming/template.Rmd \
>   /project/obds/$USER/projects/object_oriented_programming_ggplot2/object_oriented_programming-kevinrue.Rmd
> ```

]

- Similarly, make a copy of the file `template.Rmd` for the `ggplot2` lesson from the Git repository to your project.
  Use your username in the filename to make it unique, e.g.

.small-code[

> ```
> cp \
>   /project/obds/$USER/git/week2day4-ggplot2/template.Rmd \
>   /project/obds/$USER/projects/object_oriented_programming_ggplot2/ggplot2-kevinrue.Rmd
> ```

]

- Open your copy of the notebook for the object-oriented programming lesson in RStudio.

- Follow along, running code from the notebook as mentioned during the lesson.

---

```{r, child='etc.Rmd'}
```
