---
title: 'Object Oriented Programming (OOP)'
subtitle: '<i class="fab fa-r-project"></i>'
author: "Kevin Rue-Albrecht"
institute: "Oxford Biomedical Data Science Training Programme"
date: "2021-10-11 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, metropolis, rladies-fonts, "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
# uncomment this line to produce HTML and PDF in RStudio:
knit: pagedown::chrome_print
---

```{r, child='opening.Rmd'}
```

---

# Prerequisites

- A clone of the shared GitHub repository for this course.

- An account on the CCB cluster.

---

# Set up

- Pull the main branch of the shared repository.

> We have added some files to get you started.

- Make a copy of the file `solution.Rmd` for this lesson. Use your username to make it unique.

> e.g. `cp solution.Rmd solution_kevinrue.Rmd`

- Open your copy of the notebook in RStudio.

- Follow along, running code from the notebook as mentioned during the lesson.

---

```{r, child='exercises-after-setup.Rmd'}
```
