---
title: 'Empty droplets, doublets and ambient RNA'
subtitle: 'Single-cell Genomics in <i class="fab fa-r-project"></i>'
author: "Kevin Rue-Albrecht"
institute: "Oxford Biomedical Data Science Training Programme"
date: "2022-02-10 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, metropolis, rladies-fonts, "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
# uncomment this line to produce HTML and PDF in RStudio:
knit: pagedown::chrome_print
---

```{r, child='goals_and_objectives.Rmd'}
```

---

# Prerequisites

- An account on the CCB cluster.

- A clone of the shared GitHub repository for this course.

---

# Set up

- Checkout and pull the main branch of the shared repository.

- Create and launch a new RStudio project in `/project/obds/$USER/projects/emptydroplets_doublets_background`.

- Make a copy of the file `template.Rmd` for this lesson from the Git repository to your project.
  Use your username in the filename to make it unique, e.g

.small-code[

> ```
> cp \
>   /project/obds/$USER/git/week4day4-emptydroplets_doublets_background/template.Rmd \
>   /project/obds/$USER/projects/emptydroplets_doublets_background/emptydroplets_doublets_background-kevinrue.Rmd
> ```

]

- Open your copy of the notebook for this lesson in RStudio.

- Follow along, running code from the notebook as mentioned during the lesson.

The input files obtained from [1k PBMCs from a Healthy Donor (v3 chemistry)](https://www.10xgenomics.com/resources/datasets/1-k-pbm-cs-from-a-healthy-donor-v-3-chemistry-3-standard-3-0-0) are located in `/project/obds/shared/resources/4_r_single_cell/singlecell_droplets`.

---

```{r, child='etc.Rmd'}
```
