---
title: "Example code for dimensionality reduction and clustering in R"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(umap)
library(Rtsne)
library(dendextend)
library(dbscan)
```

# Exercise

## Setup

- Import the `iris` data set.

```{r}
data("iris")
```

- Separate the matrix of measurements in a new object named `iris_features`.

```{r}
iris_features <- iris %>% select(-Species)
head(iris_features)
```

# Exercise

## Apply Principal Components Analysis (PCA)

The `prcomp()` function allows you to standardise the data as part of the principal components analysis itself.

- Apply PCA while centering and scaling the matrix of features.

```{r}
pca_iris <- prcomp(iris_features, center = TRUE, scale. = TRUE)
pca_iris
```

- Examine the PCA output.
  Display the loading of each feature on each principal component.

```{r}
str(pca_iris)
```

```{r}
pca_iris$rotation
```

- Use the return value of the PCA to create a `data.frame` called `pca_iris_dataframe` that contains the coordinates projected on principal components.

```{r}
pca_iris_dataframe <- as.data.frame(pca_iris$x)
head(pca_iris_dataframe)
```

- Visualise the PCA projection using `ggplot2::geom_point()`.

```{r}
ggplot(pca_iris_dataframe, aes(PC1, PC2)) +
  geom_point() +
  cowplot::theme_cowplot()
```

### Bonus point

- Color data points according to their class label.

- Store the PCA plot as an object named `pca_iris_species`.

```{r}
pca_iris_dataframe$Species <- iris$Species
head(pca_iris_dataframe)
```

```{r}
pca_iris_species <- ggplot(pca_iris_dataframe, aes(PC1, PC2, color = Species)) +
  geom_point() +
  cowplot::theme_cowplot()
pca_iris_species
```

# Exercise

## Variable loading

- Color a scatter plot of PC1 and PC2 by the value of the variable most strongly associated with the first principal component.

What do you observe?

```{r}
pca_iris_dataframe$Petal.Width <- iris$Petal.Width
head(pca_iris_dataframe)
```

```{r}
pca_iris_petal_width <- ggplot(pca_iris_dataframe, aes(PC1, PC2, color = Petal.Width)) +
  geom_point() +
  cowplot::theme_cowplot()
pca_iris_petal_width
```

> Answer:
> - The plot reveals a gradient of the variable along PC1.
> - This is what one would expect for a variable with a high loading on that principal component.

## Variance explained

- Compute the variance explained by principal components, using information present in the return value of the `prcomp()` function.

```{r}
explained_variance_ratio <- pca_iris$sdev ^ 2 / sum(pca_iris$sdev ^ 2)
explained_variance_ratio
```

- Visualise the variance explained by each principal component using `ggplot2::geom_col()`.

```{r}
variance_dataframe <- data.frame(
  PC = seq_along(explained_variance_ratio),
  variance_ratio = explained_variance_ratio
)
head(variance_dataframe)
```

```{r}
ggplot(variance_dataframe, aes(PC, variance_ratio)) +
  geom_col() +
  ylim(0, 1) +
  cowplot::theme_cowplot()
```

# Exercise

## UMAP

- Apply UMAP on the output of the PCA.

```{r}
set.seed(1) # Set a seed for reproducible results
umap_iris <- umap(d = pca_iris$x)
umap_iris
```

- Inspect the UMAP output.

```{r}
str(umap_iris)
```

- Visualise the UMAP projection using `ggplot2::geom_point()`.

```{r}
umap_iris_dataframe <- as.data.frame(umap_iris$layout)
colnames(umap_iris_dataframe) <- paste0("UMAP", seq_len(ncol(umap_iris_dataframe)))
head(umap_iris_dataframe)
```

```{r}
ggplot(umap_iris_dataframe, aes(UMAP1, UMAP2)) +
  geom_point() +
  cowplot::theme_cowplot()
```

### Bonus point

- Color data points according to their class label.

- Store the UMAP plot as an object named `umap_iris_species`.

```{r}
umap_iris_dataframe$Species <- iris$Species
head(umap_iris_dataframe)
```

```{r}
umap_iris_species <- ggplot(umap_iris_dataframe, aes(UMAP1, UMAP2, color = Species)) +
  geom_point() +
  cowplot::theme_cowplot()
umap_iris_species
```

# Exercise

## t-SNE

- Apply t-SNE and inspect the output.

```{r}
set.seed(1) # Set a seed for reproducible results
tsne_iris <- Rtsne(pca_iris$x, check_duplicates = FALSE)
str(tsne_iris)
```

- Use the return value of the t-SNE to create a `data.frame` called `tsne_iris_dataframe` that contains the coordinates.

```{r}
tsne_iris_dataframe <- as.data.frame(tsne_iris$Y)
colnames(tsne_iris_dataframe) <- paste0("TSNE", seq_len(ncol(tsne_iris_dataframe)))
head(tsne_iris_dataframe)
```

- Visualise the t-SNE projection.

```{r}
ggplot(tsne_iris_dataframe, aes(TSNE1, TSNE2)) +
  geom_point() +
  cowplot::theme_cowplot()
```

### Bonus points

- Color data points according to their class label.

- Store the t-SNE plot as an object named `tsne_iris_species`.

```{r}
tsne_iris_dataframe$Species <- iris$Species
head(tsne_iris_dataframe)
```

```{r}
tsne_iris_species <- ggplot(tsne_iris_dataframe, aes(TSNE1, TSNE2, color = Species)) +
  geom_point() +
  cowplot::theme_cowplot()
tsne_iris_species
```

- Combine PCA, UMAP and t-SNE plots in a single figure.

```{r, fig.height=6, fig.width=6}
cowplot::plot_grid(
  pca_iris_species,
  umap_iris_species,
  tsne_iris_species,
  ncol = 2
)
```

# Exercise

## Hierarchical clustering

- Perform hierarchical clustering on the `iris_features` data set,
  using the `euclidean` distance and method `ward.D2`.
  Use the functions `dist()` and `hclust()`.

```{r}
dist_iris <- dist(iris_features, method = "euclidean")
hclust_iris_ward <- hclust(dist_iris, method = "ward.D2")
hclust_iris_ward
```

- Plot the clustering tree.
  Use the function `plot()`.

```{r}
plot(hclust_iris_ward, labels = FALSE)
```

How many clusters would you call from a visual inspection of the tree?

> Answer:
> - Visually: two major clusters.
> - One could argue for three - maybe four - clusters, depending on follow-up analyses.

- **Bonus point:** Color leaves by known species (use `dendextend`).

```{r}
iris_hclust_dend <- as.dendrogram(hclust_iris_ward)
labels_colors(iris_hclust_dend) <- as.numeric(iris$Species)
plot(iris_hclust_dend, main = "A color for every Species")
```

- Cut the tree in 3 clusters and extract the cluster label for each flower.
  Use the function `cutree()`.

```{r}
iris_hclust_ward_labels <- cutree(hclust_iris_ward, k = 3)
iris_hclust_ward_labels
```

- Repeat clustering using 3 other agglomeration methods:

  + `complete`
  + `average`
  + `single`

```{r}
# complete
hclust_iris_complete <- hclust(dist_iris, method = "complete")
iris_hclust_complete_labels <- cutree(hclust_iris_complete, k = 3)
iris_hclust_complete_labels
```

```{r}
# average
hclust_iris_average <- hclust(dist_iris, method = "average")
iris_hclust_average_labels <- cutree(hclust_iris_average, k = 3)
iris_hclust_average_labels
```

```{r}
# single
hclust_iris_single <- hclust(dist_iris, method = "single")
iris_hclust_single_labels <- cutree(hclust_iris_single, k = 3)
iris_hclust_single_labels
```

- Compare clustering results on scatter plots of the data.

```{r}
iris_clusters_dataframe <- iris
iris_clusters_dataframe$hclust_average <- as.factor(iris_hclust_average_labels)
iris_clusters_dataframe$hclust_complete <- as.factor(iris_hclust_complete_labels)
iris_clusters_dataframe$hclust_single <- as.factor(iris_hclust_single_labels)
iris_clusters_dataframe$hclust_ward <- as.factor(iris_hclust_ward_labels)
```

```{r, fig.height=8, fig.width=10}
plot_average <- ggplot(iris_clusters_dataframe, aes(Sepal.Length, Sepal.Width, color = hclust_average)) +
  geom_point() +
  cowplot::theme_cowplot()
plot_complete <- ggplot(iris_clusters_dataframe, aes(Sepal.Length, Sepal.Width, color = hclust_complete)) +
  geom_point() +
  cowplot::theme_cowplot()
plot_single <- ggplot(iris_clusters_dataframe, aes(Sepal.Length, Sepal.Width, color = hclust_single)) +
  geom_point() +
  cowplot::theme_cowplot()
plot_ward <- ggplot(iris_clusters_dataframe, aes(Sepal.Length, Sepal.Width, color = hclust_ward)) +
  geom_point() +
  cowplot::theme_cowplot()
cowplot::plot_grid(
  plot_average,
  plot_complete,
  plot_single,
  plot_ward, ncol = 2
)
```

# Exercise

## dbscan

- Apply `dbscan` to the `iris_features` data set.

```{r}
dbscan_iris <- dbscan(iris_features, eps = 0.75, minPts = 2)
dbscan_iris
```

- Visualise the `dbscan` cluster label on a scatter plot of the data.

```{r}
iris_clusters_dataframe$dbscan <- as.factor(dbscan_iris$cluster)
head(iris_clusters_dataframe)
```

```{r}
dbscan_plot <- ggplot(iris_clusters_dataframe, aes(Sepal.Length, Sepal.Width, color = dbscan)) +
  geom_point() +
  cowplot::theme_cowplot()
dbscan_plot
```

## hdbscan

- Apply `hdbscan` to the `iris_features` data set.

```{r}
hdbscan_iris <- hdbscan(iris_features, minPts = 3)
hdbscan_iris
```

- Visualise the `hdbscan` cluster label on a scatter plot of the data.

```{r}
iris_clusters_dataframe$hdbscan <- as.factor(hdbscan_iris$cluster)
head(iris_clusters_dataframe)
```

```{r}
hdbscan_plot <- ggplot(iris_clusters_dataframe, aes(Sepal.Length, Sepal.Width, color = hdbscan)) +
  geom_point() +
  cowplot::theme_cowplot()
hdbscan_plot
```

## Bonus point

- Combine the plots of `dbscan` and `hdbscan` into a single plot.

```{r, fig.height=3, fig.width=6}
cowplot::plot_grid(
  dbscan_plot,
    hdbscan_plot,
    ncol = 2
)
```

# Exercise

## K-means clustering

- Apply $K$-means clustering with $K$ set to 3 clusters.

```{r}
set.seed(1) # Set a seed for reproducible results
kmeans_iris <- kmeans(iris_features, 3)
kmeans_iris
```

- Inspect the output.

```{r}
str(kmeans_iris)
```

- Extract the cluster labels.

```{r}
kmeans_iris$cluster
```

- Extract the coordinates of the cluster centers.

```{r}
kmeans_iris$centers
```

- Construct a data frame that combines the `iris` dataset and the cluster label.

```{r}
iris_labelled <- iris
iris_labelled$Kmeans <- as.factor(kmeans_iris$cluster)
head(iris_labelled)
```

- Plot the data set as a scatter plot.

  + Color by cluster label.

```{r}
ggplot(iris_labelled, aes(Sepal.Length, Sepal.Width, color = Kmeans)) +
  geom_point() +
  cowplot::theme_cowplot()
```

### Bonus point

- Add cluster centers as points in the plot.

```{r}
iris_means_centers <- as.data.frame(kmeans_iris$centers)
iris_means_centers$Kmeans <- as.factor(seq_len(nrow(iris_means_centers)))
head(iris_means_centers)
```


```{r}
ggplot(iris_labelled, aes(Sepal.Length, Sepal.Width, color = Kmeans)) +
  geom_point() +
  geom_point(data = iris_means_centers, shape = "+", size = 20, show.legend = FALSE) +
  cowplot::theme_cowplot()
```

# Exercise

## Cross-tabulation with ground truth

- Cross-tabulate cluster labels with known labels.

```{r}
table(iris_labelled$Species, iris_labelled$Kmeans)
```

How many observations are mis-classified by $K$-means clustering?

> Answer:
> Considering that cluster 3 contains exclusively setosa observations,
> and cluster 2 contains mostly virginica observations,
> we can say that 16 observations are mis-classified:
> - 2 versicolor observations are classified with virginica observations in cluster 2
> - 14 virginica observations are classified with versicolor observations in cluster 1

## Elbow plot

- Plot the "total within-cluster sum of squares" for K ranging from 2 to 10.

```{r}
kmeans_iris$tot.withinss
```

```{r}
get_mean_totss_for_k <- function(k, data) {
  kmeans_out <- kmeans(data, k)
  return(kmeans_out$tot.withinss)
}
k_range <- 2:10
kmean_totwithinss <- vapply(k_range, get_mean_totss_for_k, FUN.VALUE = numeric(1), data = iris_features)
kmean_totwithinss
```

```{r}
kmean_totwithinss_dataframe <- data.frame(
  K = k_range,
  totss = kmean_totwithinss
)
head(kmean_totwithinss_dataframe)
```

```{r}
ggplot(kmean_totwithinss_dataframe, aes(K, totss)) +
  geom_line() +
  geom_point() +
  cowplot::theme_cowplot()
```

Do you agree that 3 is the optimal number of clusters for this data set?

> Answer:
> Without any other information than this elbow plot,
> it would be natural to claim that 4 clusters provide a reasonable compromise
> to minimise the total sum of squares within clusters
> while minimising over-clustering.
