---
title: "Template code for single-cell droplet quality control"
author: "Kevin Rue-Albrecht"
date: "2024-11-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(DropletUtils)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scDblFinder)
library(Seurat)
library(DoubletFinder)
library(celda)
library(scater)
library(sessioninfo)
```

# Exercise

## Empty droplets

- Import the raw matrix into R; use `DropletUtils::read10xCounts()`.
  Make a `SingleCellExperiment` object called `sce`.

Hints:

- Use the `samples=` argument with a named character vector,
  where the value is the path to the directory with data files,
  and the name is an arbitrary name for that sample (e.g., `pbmc5k`).
- Set the argument `col.names=` to `TRUE`,
  to use the cell barcodes as column names.

```{r}
library(DropletUtils)
sce <- DropletUtils::read10xCounts(
  samples = c("pbmc5k" = "data/raw_feature_bc_matrix"),
  col.names = TRUE
)
sce
```

- Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

Hint:

- There is no designated function for this.
  Instead, calculate the sum of UMI for each barcode in the 'counts' assay using `colSums()`,
  and use indexing with single square brackets to only keep barcodes with UMI sum greater than 0.

```{r}
sce <- sce[, colSums(assay(sce, "counts")) > 0]
sce
```

- Run `DropletUtils::emptyDrops()` on the remaining droplets.

Hint:

- Use `set.seed()` to ensure reproducibility.

```{r}
set.seed(1)
out <- DropletUtils::emptyDrops(sce)
out
```

**How do you read the output? Which droplets would you consider empty?**

> Answer:
> 
> The output is a table with a row for each GEM barcode and columns for various pieces of information.
> - `Total` represents the total UMI count.
> - `FDR` is the corrected p-value indicating the probability that a droplet is empty (i.e., originate from the null distribution).

- Remove empty droplets from the data sets.
  Call the new object `sce_filtered_emptydrops`.

Hint:

- Use the 'FDR' column of the `out` object above.
- Identify the droplets below a givne FDR threshold (e.g. 0.01).
- Use the `which()` function to get rid of NA values.
- Use indexing to keep only the desired cells from the `sce` object.

```{r}
sce_filtered_emptydrops <- sce[, which(out$FDR <= 0.01)]
sce_filtered_emptydrops
```

## Bonus point

- Draw plots comparing empty droplets and other droplets
  (e.g., library size, mitochondrial content, PCA).

```{r}
plot_data <- tibble(
  library_size = colSums(assay(sce, "counts")),
  empty_droplet = out$FDR > 0.01
)
ggplot(plot_data, aes(empty_droplet, library_size)) +
  geom_jitter(width = 0.1, height = 0) +
  scale_y_continuous(trans = "log10") +
  cowplot::theme_cowplot()
```

```{r}
is.mt <- grepl("^MT-", rowData(sce)[["Symbol"]])
# table(is.mt)
plot_data <- tibble(
  total = colSums(assay(sce, "counts")),
  MT_total = colSums(assay(sce, "counts")[is.mt, ]),
  MT_percent = MT_total / total * 100,
  empty_droplet = out$FDR > 0.01
)
ggplot(plot_data, aes(empty_droplet, MT_percent)) +
  geom_jitter(width = 0.1, height = 0) +
  #scale_y_continuous(trans = "log10") +
  cowplot::theme_cowplot()
```

```{r}
sce <- logNormCounts(sce)
sce <- runPCA(sce)
plot_data <- tibble(
  reducedDim(sce, "PCA")[, c("PC1", "PC2")] %>% as_tibble(),
  empty_droplet = out$FDR > 0.01
)
ggplot(plot_data, aes(PC1, PC2, color = empty_droplet)) +
  geom_jitter(width = 0.1, height = 0) +
  cowplot::theme_cowplot()
```

- Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- DropletUtils::read10xCounts(
  samples = c("pbmc5k" = "data/filtered_feature_bc_matrix"),
  col.names = TRUE
)
# sce_filtered_cellranger
all_barcodes <- colnames(sce)
# length(all_barcodes)
compare_table <- tibble(
  emptydrops = all_barcodes %in% colnames(sce)[out$FDR <= 0.01],
  cellranger = all_barcodes %in% colnames(sce_filtered_cellranger),
)
table(compare_table)
```

# Exercise

## Doublets

Continue using the data set after you excluded empty droplets: `sce_filtered_emptydrops`.

- Run `scDblFinder::scDblFinder()`.
  Store the output in a new object called `sce_scdblfinder`.

Hint:

- Use `set.seed()` to ensure reproducibility.

```{r}
set.seed(1)
sce_scdblfinder <- scDblFinder(sce_filtered_emptydrops)
```

- Use the `table()` function to display the number of cells annotated as doublet or singlet.

Hint:

- The `scDblFinder()` function stored that information in the colData column called `scDblFinder.class`.

```{r}
table(sce_scdblfinder$scDblFinder.class)
```

**How do you read the output? Which droplets would you consider doublets?**

> Answer:
> 
> The output is a SingleCellExperiment object.
> In the colData (cell metadata), new columns store information about each barcode.
> In particular, `sce_scdblfinder$scDblFinder.class` indicates the predicted class of each droplet: singlet or doublet.

- Remove doublets from the data set.
  Call the new object `sce_filtered_scdblfinder`.

Hint:

- Use indexing to keep only the droplets annotated as 'singlet' in the `scDblFinder.class` column of cell metadata.

```{r}
sce_filtered_scdblfinder <- sce_scdblfinder[, sce_scdblfinder$scDblFinder.class == "singlet"]
sce_filtered_scdblfinder
```

## Bonus point

- Draw plots comparing doublets and other droplets.
  (e.g., library size, mitochondrial content, PCA)

```{r}
plot_data <- tibble(
  library_size = colSums(assay(sce_scdblfinder, "counts")),
  scdblfinder = sce_scdblfinder$scDblFinder.class
)
ggplot(plot_data, aes(scdblfinder, library_size)) +
  geom_jitter(width = 0.1, height = 0) +
  scale_y_continuous(trans = "log10") +
  cowplot::theme_cowplot()
```

```{r}
is.mt <- grepl("^MT-", rowData(sce_scdblfinder)[["Symbol"]])
# table(is.mt)
plot_data <- tibble(
  total = colSums(assay(sce_scdblfinder, "counts")),
  MT_total = colSums(assay(sce_scdblfinder, "counts")[is.mt, ]),
  MT_percent = MT_total / total * 100,
  scdblfinder = sce_scdblfinder$scDblFinder.class
)
ggplot(plot_data, aes(scdblfinder, MT_percent)) +
  geom_jitter(width = 0.1, height = 0) +
  scale_y_continuous(trans = "log10") +
  cowplot::theme_cowplot()
```

```{r}
sce_scdblfinder <- logNormCounts(sce_scdblfinder)
sce_scdblfinder <- runPCA(sce_scdblfinder)
plot_data <- tibble(
  reducedDim(sce_scdblfinder, "PCA")[, c("PC1", "PC2")] %>% as_tibble(),
  scdblfinder = sce_scdblfinder$scDblFinder.class
) %>% 
  arrange(scdblfinder)
ggplot(plot_data, aes(PC1, PC2, color = scdblfinder)) +
  geom_jitter(width = 0.1, height = 0) +
  cowplot::theme_cowplot()
```

## Bonus point

- Compare results with the output of <https://github.com/chris-mcginnis-ucsf/DoubletFinder>

  + You will need to make a `Seurat` object.

```{r, message=FALSE}
## Pre-process Seurat object (standard)
seu <- CreateSeuratObject(counts = assay(sce_filtered_emptydrops, "counts"))
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu, selection.method = "vst", nfeatures = 2000)
seu <- ScaleData(seu)
seu <- RunPCA(seu)
seu <- RunUMAP(seu, dims = 1:10)
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 
oblivion <- capture.output( ## get rid of printed messages
  sweep.res.list <- paramSweep(seu, PCs = 1:10, sct = FALSE) ## still assign the value
)
sweep.stats <- summarizeSweep(sweep.res.list, GT = FALSE)
bcmvn <- find.pK(sweep.stats)
```

```{r, eval=FALSE}
ggplot(bcmvn, aes(as.numeric(as.character(pK)), BCmetric)) +
  geom_point() +
  geom_line() +
  cowplot::theme_cowplot()
```

```{r}
quick_clusters <- quickCluster(x = LayerData(object = seu, layer = "data")[VariableFeatures(seu), ])
table(quick_clusters)
```


```{r, eval=FALSE}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- modelHomotypic(quick_clusters)
nExp_poi <- round(0.01*nrow(seu@meta.data))  ## Assuming 7.5% doublet formation rate - tailor for your dataset
nExp_poi.adj <- round(nExp_poi*(1-homotypic.prop))
```

```{r, message=FALSE, eval=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- doubletFinder(seu, PCs = 1:10, pN = 0.25, pK = 0.26, nExp = nExp_poi, reuse.pANN = FALSE, sct = FALSE)
```

```{r, eval=FALSE}
table(seu$DF.classifications_0.25_0.26_53)
```

```{r, message=FALSE, eval=FALSE}
seu <- doubletFinder(seu, PCs = 1:10, pN = 0.25, pK = 0.26, nExp = nExp_poi.adj, reuse.pANN = "DF.classifications_0.25_0.26_53", sct = FALSE)
```

```{r, eval=FALSE}
table(seu$DF.classifications_0.25_0.26_48)
```

```{r, eval=FALSE}
doublets_table <- data.frame(
  row.names = colnames(sce_filtered_emptydrops),
  scDblFinder = factor(sce_scdblfinder$scDblFinder.class, c("singlet", "doublet")),
  DoubletFinder = factor(seu$DF.classifications_0.25_0.26_53, c("Singlet", "Doublet"))
)
table(doublets_table)
```

# Exercise

## Ambient RNA

Continue using the data set after excluding empty droplets and doublets: `sce_filtered_scdblfinder`.

- Run `celda::decontX()`.

Hint:

- Use `set.seed()` to ensure reproducibility.

```{r, message=FALSE}
set.seed(1)
sce_decontX <- celda::decontX(sce_filtered_scdblfinder)
```

- Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.

Hint:

- Look at the example section of the help page.
  You will find that the authors use the arguments `x=`, `dim1=` and `dim2=`.

```{r}
umap <- reducedDim(sce_decontX, "decontX_UMAP")
plotDimReduceCluster(x = sce_decontX$decontX_clusters, dim1 = umap[, 1], dim2 = umap[, 2])
```

- Visualize the percentage of contamination estimated by `decontX` in each cell.
  Use the `plotDecontXContamination()` function.

```{r}
plotDecontXContamination(sce_decontX)
```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer:
> 
> Cells in cluster 5 (see previous plot) show a highlevel of contamination (close to 1).

- Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->

Hint:

- Look at the documentation; you will see that developers provide a matrix of log-normalised counts,
  and use the arguments `dim1=` and `dim2=`.
- First, use the function `logNormCounts()` to add the matrix of log-normalised 'raw' counts to the object.
- At this point, genes are still identified by their Ensembl identifier.
  Here, we use the function `uniquifyFeatureNames()` to identify genes by a unique combination of gene symbol and Ensembl identifier.
- Use the argument `exactMatch=` to indicate that you are giving exact feature identifiers.

```{r}
sce_decontX <- logNormCounts(sce_decontX, assay.type = "counts")
rownames(sce_decontX) <- uniquifyFeatureNames(
  ID = rownames(sce_decontX),
  names = rowData(sce_decontX)[["Symbol"]]
)
plotDimReduceFeature(
  as.matrix(logcounts(sce_decontX)),
  dim1 = umap[, 1],
  dim2 = umap[, 2],
  features = c("CD3D", "CD3E", "GNLY",
  "LYZ", "S100A8", "S100A9",
  "CD79A", "CD79B", "MS4A1"),
  exactMatch = TRUE
)
```

- Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.

Hint:

- First, make a (base R) list of markers, grouped by cell type.
- Then, make a (base R) list of cell types of interest, indicating the cluster(s)
  that contain those cell types.
- Finally, provide those two lists to the arguments `markers=` and `groupClusters=`
  of the function `plotDecontXMarkerPercentage()`.
- The multiple code chunks below are here for you
  to run the `plotDecontXMarkerPercentage()` function multiple times,
  either on the assay 'counts', 'decontXcounts', or both.

```{r}
markers <- list(
  Tcell_Markers = c("CD3E", "CD3D"),
  Bcell_Markers = c("CD79A", "CD79B", "MS4A1"),
  Monocyte_Markers = c("S100A8", "S100A9", "LYZ"),
  NKcell_Markers = "GNLY"
)
cellTypeMappings <- list(
  Bcells = 4,
  Monocytes = 2,
  Tcells = c(1, 3),
  NKcells = c(5, 6)
)
plotDecontXMarkerPercentage(
  sce_decontX,
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = "counts"
)
```

```{r}
plotDecontXMarkerPercentage(
  sce_decontX,
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = "decontXcounts"
)
```

```{r}
plotDecontXMarkerPercentage(
  sce_decontX,
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = c("counts", "decontXcounts")
)
```

- Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

Hint:

- This function produces a facet for each marker in each cell type.
  That can make a lot of facets!
  To limit the number of facets, consider giving the function the markers for just one cell type
  e.g. `markers[["Monocyte_Markers"]]`.
- Use the argument `groupClusters=` to provide the list of cell types and associated clusters defined earlier.

```{r}
plotDecontXMarkerExpression(
  sce_decontX,
  markers = markers[["Monocyte_Markers"]],
  groupClusters = cellTypeMappings,
  ncol = 3
)
```

# Session info

```{r}
session_info()
```
