---
title: "Solution code for tidyverse"
author: "Kevin Rue-Albrecht"
date: "18/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Demo

- Load the tidyverse

```{r}
library(tidyverse)
```

- Load a single package from the tidyverse

```{r}
library(ggplot2)
```

# Demo

# The pipe operator

```{r}
x <- 4
x %>% sqrt()
```

```{r}
x <- 4
sqrt(x)
```

```{r}
iris %>% 
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point()
```

# Demo

## The tidyverse philosophy

```{r}
iris %>%
    select(Sepal.Length, Sepal.Width, Species) %>%
    slice_head(n = 3)
```

# Exercise

## Read and write files

- Read data from the file `iris.csv`.
  Assign the data imported from the file to an object called `iris_raw`.

```{r}
iris_raw <- read_csv("data/iris.csv")
```

**What do you learn about the data from the messages displayed in the R console while the contents of the file are parsed and imported into your R session?**

> Answer:
> - Number of rows and columns in the data set.
> - Delimiter used to parse the file.
> - Name and type of individual columns.
> - Hints about some functions and options that can be used.

- Print the value of `iris_raw`.

```{r}
iris_raw
```

**What is the class of the data? What are the dimensions of the dataset? What is the type of data stored in each column?**

> Answer:
> - Data are loaded as a `tibble`.
> - Data set is 150 rows and 5 columns.
> - Column types are: double, double, double, double, character.

- Write the dataset to a file named `iris.tsv`, separating fields with the tabulation character.

```{r}
write_tsv(iris_raw[, 1:4], "iris.tsv")
```

**What function do you use? What options are available for that function?**

> Answer:
> - `write.tsv()` is the ideal choice here.
> - Options can be found on the man page, using `?write.tsv`.

- Inspect the `iris.tsv` file.
  You can use `file.edit()` to open the file in the RStudio editor.

```{r, eval=FALSE}
file.edit("iris.tsv")
```

**Are you satisfied with the contents and appearance of the file?**

> Answer:
> - The file is as expected.
> - Not the nicest to read, due to column names wider than values.

# Demo

## Making a tibble

```{r}
tibble(x = 1:5, y = 1, z = x ^ 2 + y)
```

# Demo

## Subset the columns of a table

```{r}
iris %>%
    select(Sepal.Length, Sepal.Width) %>% 
    slice_head(n = 6)
```

```{r}
iris %>%
    select(starts_with("Petal") | ends_with("Width")) %>% 
    slice_head(n = 6)
```

```{r}
iris %>%
    select(!ends_with("Width")) %>% 
    slice_head(n = 6)
```

```{r}
iris %>%
    select(!c(Sepal.Length, Petal.Length)) %>% 
    slice_head(n = 6)
```

# Demo

## Create and update columns in a table

```{r}
iris %>%
    mutate(
        ID = seq(1, nrow(iris)),
        Flower.ID = paste0(Species, ID)
        ) %>%
    slice_head(n = 6)
```

# Demo

## Subset observations in a table

```{r}
iris %>%
    filter(Sepal.Length > mean(Sepal.Length) & Sepal.Width > mean(Sepal.Width)) %>%
    as_tibble()
```

# Demo

## Compute summary statistics

Without grouping

```{r}
iris %>%
    summarise(Sepal.Length.mean = mean(Sepal.Length))
```

With grouping

```{r}
iris %>%
    group_by(Species) %>%
    summarise(Sepal.Length.mean = mean(Sepal.Length))
```

# Demo

## Sort observations

```{r}
iris %>%
    arrange(Species, desc(Sepal.Length)) %>%
    as_tibble()
```

# Demo

## Extract a single column as a vector

Without names

```{r}
iris %>%
    pull(Sepal.Length) %>%
    head(5)
```

With names

```{r}
iris %>%
    pull(Sepal.Length, name = Species) %>%
    head(5)
```

# Demo

## Combine two tables using shared information

```{r}
tibble_1 <- tibble(
  ID = paste0("sample", 1:4),
  gene1 = rbinom(4, 10, 0.5),
  gene2 = rbinom(4, 10, 0.5)
)
tibble_1
```

```{r}
tibble_2 <- tibble(
  ID = paste0("sample", 1:4),
  batch = factor(rep(c("A", "B"), each = 2)),
  condition = factor(rep(c("control", "treated"), times = 2)),
)
tibble_2
```

**How would you describe how to join these two tibbles?**

```{r}
tibble_joined <- left_join(tibble_1, tibble_2, by = "ID")
tibble_joined
```

# Exercise

## Manipulate data

### Exercise 1

- Using `iris_raw`, for each species of iris, compute the following summary statistics for the `sepal_length`: mean, median, minimum, maximum.

```{r}
iris_raw %>% 
    group_by(species) %>% 
    summarise(
        min = min(sepal_length),
        mean = mean(sepal_length),
        median = median(sepal_length),
        max = max(sepal_length)
    )
```

### Exercise 2

- For each species of iris, compute the mean of every column that is numeric.
  **Hint:** use the functions `dplyr::across()`, `tidyselect::where()`, and `base::is.numeric()`.

```{r}
iris_raw %>% 
    group_by(species) %>% 
    summarise(across(where(is.numeric), mean))
```

- Filter the table above to retain only species of iris with an average sepal length less than `6`.

```{r}
# Copy the code chunk above and extend with more pipes
iris_raw %>% 
    group_by(species) %>% 
    summarise(across(tidyselect::starts_with("sepal_"), mean)) %>% 
    filter(sepal_length < 6)
```

- Sort the table above by descending `sepal_length`.

```{r}
# Copy the code chunk above and extend with more pipes
iris_raw %>% 
    group_by(species) %>% 
    summarise(across(tidyselect::starts_with("sepal_"), mean)) %>% 
    filter(sepal_length < 6) %>% 
    arrange(desc(sepal_length))
```

- From the table above, extract the `sepal_length` column as a numeric vector.
  Make it a named numeric vector, where each value is named with the corresponding species.

```{r}
# Copy the code chunk above and extend with more pipes
iris_raw %>% 
    group_by(species) %>% 
    summarise(across(tidyselect::starts_with("sepal_"), mean)) %>% 
    filter(sepal_length < 6) %>% 
    arrange(desc(sepal_length)) %>% 
    pull(sepal_length, name = species)
```

# Exercise

## Manipulate data

### Exercise 3

Let's make the silly assumption that iris sepals are rectangular in shape.

- Using `iris_raw`, compute a new column named `sepal_area`, which is the product of `sepal_length` and `sepal_width`.

```{r}
iris_raw %>% 
    mutate(
        sepal_area = sepal_length * sepal_width
    )
```

- Subset the result to the columns named `species` and `sepal_area`.

```{r}
# Copy the code chunk above and extend with more pipes
iris_raw %>% 
    mutate(
        sepal_area = sepal_length * sepal_width
    ) %>% 
    select(species, sepal_area)
```

- Subset the result to display the top 5 observations by `sepal_area`.

```{r}
# Copy the code chunk above and extend with more pipes
iris_raw %>% 
    mutate(
        sepal_area = sepal_length * sepal_width
    ) %>% 
    select(species, sepal_area) %>% 
    slice_max(sepal_area, n = 5)
```

## Bonus point

- Make a histogram of `sepal_area` colored by species.

You might also want to facet the plot by species.

```{r}
# Copy the code chunk above and extend with more pipes
iris_for_ggplot <- iris_raw %>% 
    mutate(
        sepal_area = sepal_length * sepal_width
    ) %>% 
    select(species, sepal_area)
iris_for_ggplot
```

```{r}
ggplot(iris_for_ggplot, aes(sepal_area, fill = species)) +
    geom_histogram() +
    facet_wrap(~species, ncol = 1)
```

# Exercise

## Pivot data from wide to long

Reshape the `iris_raw` dataset in a tidy format where one observation is represented by:

- the species

- the variable measured

- the value

**Hint:** you want to pivot all the columns that start are numeric.

```{r}
iris_long <- iris_raw %>% 
  pivot_longer(where(is.numeric), names_to = "variable")
iris_long
```

**What information have we lost in the process? What could we do to remedy the issue?**

> Answer:
> The identity of each observation is lost;
> e.g., we cannot tell which value of `sepal_length` is associated with which value of `sepal_width`.

```{r}
# Copy the code chunk above and refine to address the issue
iris_long2 <- iris_raw %>% 
  rownames_to_column("specimen") %>% 
  pivot_longer(c(starts_with("Sepal"), starts_with("Petal")), names_to = "variable")
iris_long2
```

# Exercise

## Pivot data from long to wide

- Reshape the tidy format of the iris data set into the original wide format.

**Hint:** you will only be able to restore the wide format if you kept track of the identity of each flower in the long format.

```{r}
iris_long2 %>% 
      pivot_wider(id_cols = c(specimen, species), names_from = variable, values_from = value) %>% 
  select(-specimen)
```

# Demo

## Split a column value into multiple columns

```{r}
iris %>% 
    separate(Sepal.Length, c("Sepal.Length.unit", "Sepal.Length.decimal"), sep = "[.]") %>%
    select(c("Sepal.Length.unit", "Sepal.Length.decimal")) %>%
    as_tibble()
```

# Demo

## Combine multiple columns into a single value

```{r}
iris %>% 
  mutate(ID = seq(1, nrow(iris))) %>% 
  unite("FlowerID", Species, ID, sep = "_") %>% 
  as_tibble()
```

# Demo

## Extract substrings

```{r}
iris_species <- iris %>%
    pull(Species)
```

```{r}
iris_species %>%
    str_sub(1, 3) %>%
    unique()
```

```{r}
str_sub(iris_species, 4) <- "..."
iris_species %>%
    unique()
```

# Demo

## Join multiple strings and remove whitespaces

```{r}
words <- c("A ", " few ", "words")
words
```

```{r}
words %>%
    str_trim()
```

```{r}
words %>%
    str_trim() %>%
    str_c(collapse = " ")
```

# Integrated exercise

## Gene expression

### Part 1

We provide you with an Excel file that contains gene expression data for the nine most highly variable genes in a pancreas cancer gene expression dataset (<https://portal.gdc.cancer.gov/projects/TCGA-PAAD>), as well as metadata for the individual samples and genes.

- Import the gene expression matrix from the Excel file into the R session as a `tibble`.
  Assign it to an object named `exprs_assay`.

```{r}
library(readxl)
excel_sheets("data/EH2011.xlsx")
exprs_assay <- read_xlsx("data/EH2011.xlsx", "exprs")
exprs_assay
```

- Reshape the gene expression matrix in a tidy format.
  Assign it to a new object `exprs_assay_long`.

```{r}
exprs_assay_long <- exprs_assay %>%
  pivot_longer(-gene, names_to = "sample", values_to = "exprs")
exprs_assay_long
```

- Draw a histogram of gene expression levels for each gene in a separate facet.

```{r}
ggplot(exprs_assay_long, aes(exprs)) +
  geom_histogram(fill = "grey", color = "black", bins = 40) +
  facet_wrap(~gene) +
  cowplot::theme_cowplot()
```

---

# Integrated exercise

## Gene expression

### Part 2

- Import the sample metadata from the Excel file into the R session as a `tibble`.
  Assign it to an object named `sample_info`.

```{r}
excel_sheets("data/EH2011.xlsx")
sample_info <- read_xlsx("data/EH2011.xlsx", "sample_info")
sample_info
```

> Metadata key:
> 
> [NIH AJCC staging system](https://www.cancer.gov/publications/dictionaries/ca> ncer-terms/def/ajcc-staging-system)
> 
> - grade: how abnormal the cancer cells and tissue look under a microscope
> - T: size of the tumor and any spread of cancer into nearby tissue
> - N: spread of cancer to nearby lymph nodes
> - M: metastasis (spread of cancer to other parts of the body)

- Convert the four columns above to the factor type.

```{r}
sample_info <- sample_info %>%
  mutate(across(
    c(sex, grade, T, N, M, vital_status, sample_type),
    list(as.factor),
    .names = "{.col}"))
sample_info
```

- Combine sample information with gene expression data.
  Assign the output to `exprs_sampleinfo`.

```{r}
exprs_sampleinfo <- sample_info %>%
  left_join(exprs_assay_long, c("unique_patient_ID" = "sample"))
exprs_sampleinfo
```

---

# Integrated exercise

## Gene expression

### Part 3

- Identify the gene with the largest absolute difference in expression between male and female patients.

```{r}
exprs_sampleinfo %>% 
  group_by(sex, gene) %>% 
  summarise(exprs = mean(exprs)) %>% 
  pivot_wider(names_from = sex, values_from = exprs) %>% 
  mutate(F_minus_M = F - M) %>% 
  arrange(desc(abs(F_minus_M)))
```

- Visualise the expression of that gene across tumour grades.
  Facet by `sex`.
  Color by `T` score.

```{r}
table_for_ggplot <- exprs_sampleinfo %>% 
  filter(gene == "XIST")
table_for_ggplot
```

```{r}
ggplot(table_for_ggplot, aes(grade, exprs, colour = T)) +
  geom_point() +
  facet_wrap(~ sex)
```

- Draw a heat map of all nine genes.
  Remove labels as well as ticks for the x-axis.
  Remove the panel grid and background.

```{r}
ggplot(exprs_sampleinfo, aes(unique_patient_ID, gene, fill = exprs)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme(
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    panel.grid = element_blank(), panel.background = element_blank())
```

### Bonus point

- In the heat map, group samples by sex.

```{r}
# Copy the code chunk above and tweak the aes(...) contents.
ggplot(exprs_sampleinfo, aes(interaction(unique_patient_ID, sex), gene, fill = exprs)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme(
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
    panel.grid = element_blank(), panel.background = element_blank())
```
