---
title: 'Introduction to `r BiocStyle::CRANpkg("ggplot2")`'
subtitle: '<i class="fab fa-r-project"></i>'
author: "Kevin Rue-Albrecht"
institute: "Oxford Biomedical Data Science Training Programme"
date: "2022-05-12 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, metropolis, rladies-fonts, "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
# uncomment this line to produce HTML and PDF in RStudio:
knit: pagedown::chrome_print
---

```{r, child='goals_and_objectives.Rmd'}
```

---

# Prerequisites

- An account on the CCB cluster.

- A clone of the shared GitHub repository for this course.

---

# Set up

- Launch RStudio Server on the teaching cluster.

- Start a new RStudio project called `ggplot2`.

- Make a copy of the template notebook for this lesson in your <i class="fab fa-git"></i> repository.

- Make symbolic link from the RStudio project to the template notebook.

- Open the notebook and follow along, editing and running the code as needed.

---

```{r, child='etc.Rmd'}
```
