---
title: 'Integrated tidyverse exercise'
subtitle: '<i class="fab fa-r-project"></i> on the CCB cluster'
author: "Kevin Rue-Albrecht"
institute: "Oxford Biomedical Data Science Training Programme"
date: "2023-06-09 (updated: `r Sys.Date()`)"
date-format: long
format:
  revealjs: 
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    logo: img/ox_brand1_pos.gif
    css: styles.css
    fontsize: 1.5em
    footer: Kevin Rue-Albrecht -- Integrated tidyverse exercise
# run the following line after knitting to convert HTML to PDF
# renderthis::to_pdf("index.cluster.html")
---

```{r setup, include = FALSE}
stopifnot(requireNamespace("htmltools"))
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE, error = FALSE,
  include = FALSE
)
stopifnot(require(magrittr))
```

```{r, load_refs, include=FALSE, cache=FALSE}
options(htmltools.dir.version = FALSE)
library(RefManageR)
BibOptions(
  check.entries = FALSE,
  bib.style = "authoryear",
  cite.style = "authoryear",
  max.names = 2,
  style = "markdown",
  hyperlink = "to.doc",
  dashed = TRUE)
bib <- ReadBib("bibliography.bib")
NoCite(bib, "R")
```

```{r, child='goals_and_objectives.qmd'}
```

<!--
Docs:
- https://quarto.org/docs/get-started/hello/rstudio.html
- https://quarto.org/docs/presentations/revealjs/
- https://github.com/quarto-dev/quarto-web/blob/main/docs/presentations/revealjs/demo/index.qmd
- https://quarto.org/docs/reference/formats/presentations/revealjs.html#slides
- https://quarto.org/docs/output-formats/html-themes.html
-->

## Prerequisites

- An account on the teaching cluster.

- A [<i class="fab fa-github"></i> GitHub](https://github.com/) repository for this course.

## Set up

- Launch RStudio Server on the teaching cluster.

- Start a new RStudio project called `tidyverse_integrated`.

- Make a copy of the template notebook for this lesson in your <i class="fab fa-git"></i> repository.

- Make a symbolic link from the RStudio project to the template notebook.

- Open the notebook and follow along, editing and running the code as needed.

```{r, include=TRUE, echo=FALSE, fig.align='center', out.height='250px'}
## Source: Adobe Illustrator (Kevin Rue-Albrecht)
knitr::include_graphics("img/obds-git-setup.svg")
```

```{r, child='etc.qmd'}
```
