---
title: 'Building a tidyverse workflow'
author: "Kevin Rue-Albrecht"
date: "2024-10-16"
date-format: long
format: html
---

```{r}
#| label: load-packages
#| include: false

library(tidyverse)
library(readxl)
```

## Exercise

### Part 1

We provide you with an Excel file `EH2011.xlsx` that contains gene expression data for the nine most highly variable genes in a pancreas cancer gene expression dataset (downloaded from <https://portal.gdc.cancer.gov/projects/TCGA-PAAD>), as well as metadata for the individual samples and genes.

- Import the gene expression matrix from the Excel file into the R session as a `tibble`.
  Assign it to an object named `exprs_assay`.

```{r}
library(   )
excel_sheets(   )
exprs_assay <- 
exprs_assay
```

- Reshape the gene expression matrix in a tidy format.
  Assign it to a new object `exprs_assay_long`.

```{r}
exprs_assay_long <- exprs_assay %>%

exprs_assay_long
```

- Draw a histogram of gene expression levels for each gene in a separate facet.

```{r}
ggplot(exprs_assay_long,   ) +



```

## Exercise

### Part 2

- Import the sample metadata from the Excel file into the R session as a `tibble`.
  Assign it to an object named `sample_info`.

```{r}
excel_sheets(   )
sample_info <- 
sample_info
```

> Metadata key:
> 
> [NIH AJCC staging system](https://www.cancer.gov/about-cancer/diagnosis-staging/staging)
> 
> - sex: male or female
> - grade: how abnormal the cancer cells and tissue look under a microscope
> - T: size of the tumor and any spread of cancer into nearby tissue
> - N: spread of cancer to nearby lymph nodes
> - M: metastasis (spread of cancer to other parts of the body)
> - vital_status: living or deceased
> - sample_type: tumour or normal

- Convert each of the columns above to factors.

```{r}
sample_info <- sample_info %>%




sample_info
```

- Combine sample information with gene expression data.
  Assign the output to `exprs_sampleinfo`.

```{r}
exprs_sampleinfo <- sample_info %>%

exprs_sampleinfo
```

## Exercise

### Part 3

- Identify the gene with the largest absolute difference in expression between male and female patients.

```{r}
exprs_sampleinfo %>% 





```

- Visualise the expression of that gene in the two sexes using a geom of your choice.
  Consider colouring and faceting by other metadata (e.g., `grade` and `T`).

```{r}
table_for_ggplot <- exprs_sampleinfo %>% 

table_for_ggplot
```

```{r}
ggplot(table_for_ggplot,   ) +


```

- Draw a heat map of all nine genes.
  Remove labels as well as ticks for the x-axis.
  Remove the panel grid and background.

```{r}
ggplot(exprs_sampleinfo,   ) +





```

- Repeat the heat map, faceting samples by sex.

```{r}
# Copy the code chunk above and edit as needed
ggplot(exprs_sampleinfo,   ) +





```
