---
title: 'Introduction to base <i class="fab fa-r-project"></i>'
subtitle: 'Introduction to <i class="fab fa-r-project"></i>'
author: "Kevin Rue-Albrecht"
institute: "Oxford Biomedical Data Science Training Programme"
date: "2024-02-19 (updated: `r Sys.Date()`)"
date-format: long
format:
  revealjs: 
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    logo: img/ox_brand1_pos.gif
    css: styles.css
    fontsize: 1.5em
    footer: Kevin Rue-Albrecht -- Introduction to base <i class="fab fa-r-project"></i>
# run the following line after knitting to convert HTML to PDF
# renderthis::to_pdf("index.cluster.html")
---

```{r}
#| label: setup
#| include: false
stopifnot(requireNamespace("htmltools"))
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
options(width = 120)
```

```{r}
#| label: load_refs
#| include: false
#| cache: false
options(htmltools.dir.version = FALSE)
library(RefManageR)
BibOptions(
  check.entries = FALSE,
  bib.style = "authoryear",
  cite.style = "authoryear",
  max.names = 2,
  style = "markdown",
  hyperlink = "to.doc",
  dashed = TRUE)
bib <- ReadBib("bibliography.bib")
NoCite(bib, "R")
```

## Credits

- Based on slides by Nicki Gray.
- Liezel Tamon

```{r}
#| child: goals_and_objectives.qmd
```

## Prerequisites

- An account on the teaching cluster.

## Set up

- Open OpenVPN Connect and connect to the WIMM VPN using your WIMM IT username and password, followed by the 2FA code generated by your authenticator app.
- Open a new tab in your web browser and navigate to <https://obds.imm.ox.ac.uk> to use RStudio in the teaching server.
- Type your WIMM IT username (SSO) and password.

```{r}
#| child: etc.qmd
```
